---
title: "Final Rough Draft"
author: "Blake Behrens"
date: "2024-11-04"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rvest)
library(tidyverse)
library(lubridate)
library(reshape2)
library(gridExtra)
```

# Proposal:

In this project, we will look at team statistics from Big 12 Football teams. We will look at what variable correlate most with win percentage, analyze which teams are best in which categories, and see which teams are most efficient. \n

### Data Collection
The first thing we must do is create a Data Set for each of the Big 12 Football Teams. To do this, I will use a technique called 'webscraping', where I access each team's website on ESPN and pull the important data from the website's HTML:
```{r urls}
urls = c('https://www.espn.com/college-football/team/stats/_/type/team/id/66',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/252',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/38',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/2306',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/2641',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/2132',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/9',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/277',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/239',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/248',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/2628',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/2116',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/254',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/2305',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/12',
         'https://www.espn.com/college-football/team/stats/_/type/team/id/197')
ids = c('Iowa State','BYU','Colorado','Kansas State','Texas Tech','Cincinnati','Arizona State','West Virginia','Baylor','Houston','TCU','UCF','Utah','Kansas','Arizona','Oklahoma State')
length(ids) == length(urls)
```
This code segment identifies the ESPN link to each team's team statistics. Then, I give an `id` to each link to identify which team it belongs to, and check that the length of both variables is the same<br>

Now, lets write a base case for one url:
```{r base case}
site = read_html(urls[1])
tbl = site %>% html_nodes(".Table__TD") %>% html_text()
df = data.frame("Category" = tbl[1:50], "Stat" = tbl[51:100])
df$team = ids[1]
head(df)
str(df)
```
Now we can see that this is the data we want to see for `Iowa State`, lets apply it to each team and create a larger data frame with each team's information:
```{r create df}
i=1
large.df = data.frame()
for(url in urls){
  #Access url's html
  site = read_html(url)
  #Scrape wanted data
  tbl = site %>% html_nodes(".Table__TD") %>% html_text()
  record = site %>% html_nodes(".ml4 li:nth-child(1)") %>% html_text()
  #organize into data frame
  df = data.frame("Category" = tbl[1:50], "Stat" = tbl[51:100])
  #label data, and increase iteration
  df$Team = ids[i]
  df$record = record
  i = i +1
  #combine small with large dataset
  large.df = rbind(large.df,df)
}
```
We now have our large dataset that we desired, but its not formatted very well. The first thing we want to do is remove rows where the category is a description, and not a statistic. We can find these rows easily because they do not have a value in the `Stat` column. 
```{r filter}
df = large.df %>% 
  filter(Stat != '')
head(df)
```
Next, I want to make each team have a row, with each statistic being a column. Its also important to rename some categories so their names do not begin with a number
```{r organize}
df = df %>% 
  pivot_wider(names_from = Category, values_from = Stat)
colnames(df)[9:13] = c('first downs by penalty','third down efficiency','third down percentage','fourth down efficiency','fourth down percentage')
head(df)
str(df)

```


### Questions to Be Answered: \n
Now that we have a dataset, what questions do we want to answer:

1. What variables correlate best with winning percentage?
2. What are different ways to measure offensive efficiency?
3. Who are the most efficient offensive teams in the Big 12?
4. Which team is best in each Statistical Category?<br>


### More Organization
Before we can answer those questions, lets do some further organization of our data.<br>
First, we want to make sure all columns are formatted as single numbers, not combinations. Meaning, columns like `Sacks - Yards lost` should be turned into multiple columns, `Sacks` and `Yards Lost`<br>
```{r separation}
df = df %>% 
  separate(`third down efficiency`, '-', 
           into = c('third.converted','third.attempted') ) %>% 
  separate(`fourth down efficiency`, '-', 
           into = c('fourth.converted','fourth.attempted') ) %>% 
  separate(`Comp-Att`, '-', 
           into = c('completions','pass.attempts') ) %>% 
  separate(`Sacks-Yards Lost`, '--', 
           into = c('sacks.lost','yds.lost.sacks') ) %>% 
  separate(`Kickoffs: Total`, '-', 
           into = c('kickoffs.returned','kickoff.yards.returned') ) %>% 
  separate(`Punt: Total`, '-', 
           into = c('punts.returned','punt.yards.returned') ) %>% 
  separate(`INT: Total`, '-', 
           into = c('interceptions','int.yards.returned') ) %>% 
  separate(`Punt: Total Yards`, '-', 
           into = c('punts','punt.yards') ) %>% 
  separate(`FG: Good-Attempts`, '-', 
           into = c('FG.made','fg.attempts') ) %>% 
  separate(`Total-Yards`, '-', 
           into = c('penalties.against','penalized.yards') ) %>% 
  separate(`Fumbles-Lost`, '-', 
           into = c('fumbles','fumbles.lost') ) %>% 
  separate(record, '-',
             into = c('wins','losses')) %>% 
  mutate(t.o.p = (as.numeric(substr(`Possession Time Seconds`,1,2)) * 60) + as.numeric(substr(`Possession Time Seconds`,4,5)))
  
```

Now, before we go any further, since each column, other than `Team`, is numeric we must covert them to be a numeric data type. In order to do that, we must also re-format the column names so there is no whitespace. Furthermore, remove any commas from numbers that are in the thousands
```{r mutate data}
new.names = gsub(' ','.', colnames(df))
colnames(df) = new.names
#Remove any commas in the entire data frame
df[] <- lapply(df, function(x) gsub(",", "", as.character(x)))

#Make columns 2:n numeric
for (col in colnames(df)[2:ncol(df)]){
  df[[col]] = as.numeric(df[[col]])
}
```

Our data is now accessible as numeric values, so lets look at which columns we will be interested in. We must also add `games` as the sum of `wins` and `losses`. Additionally, we summarize some statistics like `completion.rate` instead of total `completions`.
```{r mutate pt 2}
df = df %>% 
  mutate(games = wins + losses) %>% 
  select(Team, wins, losses, games, Total.Points.Per.Game,
         Total.Touchdowns,Total.1st.downs,Rushing.1st.downs,Passing.1st.downs,
         first.downs.by.penalty,third.down.percentage,fourth.down.percentage,
         completions,pass.attempts,Yards.Per.Pass.Attempt,Passing.Touchdowns,
         Interceptions,sacks.lost,yds.lost.sacks,Yards.Per.Rush.Attempt,
         Rushing.Yards.Per.Game,Rushing.Touchdowns,Total.Offensive.Plays
         ,Yards.Per.Game,interceptions,FG.made,fg.attempts,penalties.against,
         Turnover.Ratio,t.o.p
         ) %>% 
  mutate(completion.rate = completions/pass.attempts) %>% 
  mutate(fg.percentage = FG.made / fg.attempts) %>% 
  select(-c(completions,pass.attempts,FG.made,fg.attempts))
```
A good rule of thumb in sports statistics is to use stats as an 'average per game' value instead of total, in case teams play different number of games in a season. With that in mind, I will convert certain columns to this format.
```{r avg per game}
df = df %>% 
  mutate(td.pg = Total.Touchdowns / games) %>% 
  mutate(first.pg = Total.1st.downs / games) %>% 
  mutate(rush.1st.prop = Rushing.1st.downs / Total.1st.downs) %>% 
  mutate(pass.1st.prop = Passing.1st.downs / Total.1st.downs) %>% 
  mutate(pTD.pg = Passing.Touchdowns / games) %>% 
  mutate(int.thrown.pg = Interceptions / games) %>% 
  mutate(sacks.l.pg = sacks.lost / games) %>% 
  mutate(yds.persack = yds.lost.sacks / sacks.lost) %>% 
  mutate(rTD.pg = Rushing.Touchdowns / games) %>% 
  mutate(off.plays.pg = Total.Offensive.Plays / games) %>% 
  mutate(def.int.pg = interceptions / games) %>% 
  mutate(penal.pg = penalties.against / games) %>% 
  mutate(t.o.p = t.o.p / 3600) %>% 
  mutate(wp = wins / games)
```
Now we have a **lot** of columns. Lets go through and select the ones we don't need. However, I will rename the data frame, in case we ever want to access these generalized statistics again.
```{r clean columns}
data = df %>% 
  select(-c(Total.Touchdowns, Total.1st.downs,Rushing.1st.downs,
            Passing.1st.downs,Passing.Touchdowns,Interceptions,
            sacks.lost,yds.lost.sacks,Rushing.Touchdowns,
            Total.Offensive.Plays,interceptions,penalties.against,wins,losses))
```

### Answering Questions
#### Correlations
First, lets look at the correlations between each column and win percentage:<br>
```{r correlation}
# All correlations
correlations = data.frame(cor(data[,3:ncol(data)]))

# Look at only the correlations compare to WP
correlations = correlations %>% 
  select(wp) %>% 
  arrange(-abs(wp)) %>% 
  t() %>%
  data.frame() %>% 
  select(-wp)%>% 
  t() %>%
  data.frame()
```
<br> Step by step, this finds the correlations of all variables to one another. Then, it selects the correlations between win percentages and all other variables, and arranges them based on the magnitude of the correlation. Lastly, we maninpulate the data slightly to format it in a readable way.<br>

Now, lets look at the most important variables that correlate with win percentages:<br>
```{r top correlations}
top_correlations = head(correlations, 6)
top_correlations
```
<br> For all teams, we can see that the top 2 variables that correlate most with win percentage is `Turnover.Ratio` and `int.thrown.pg` (interceptions thrown per game). The next few variables have to do with offensive efficiency, with variables such as `Total.Points.Per.Game`, `Yards.Per.Pass.Attempt`, `td.pg` and `pTd.pg` (touchdowns per game, and passing touchdowns per game, respectively).<br>

<br> Let's plot those variables against win percentage to see their correlations<br>
```{r}
p1 = ggplot(data, aes(x=Turnover.Ratio,y=wp))+
  geom_point(color = 'red')+
  geom_smooth(method='lm', se = F, color = 'black')+
  xlab('Turnover Ratio')+
  ylab("Win Percentage")
p2 = ggplot(data, aes(x = int.thrown.pg, y = wp)) + 
  geom_point(color = 'blue')+
  geom_smooth(method='lm', se = F, color = 'black')+
  xlab('Interceptions Thrown Per Game')+
  ylab("Win Percentage")
p3 = ggplot(data, aes(x = Total.Points.Per.Game, y = wp)) + 
  geom_point(color = 'black')+
  geom_smooth(method='lm', se = F, color = 'black')+
  xlab('Points Per Game')+
  ylab("Win Percentage")
p4 = ggplot(data, aes(x = Yards.Per.Pass.Attempt, y = wp)) + 
  geom_point(color = 'orange')+
  geom_smooth(method='lm', se = F, color = 'black')+
  xlab('Yards per Pass Attempt')+
  ylab("Win Percentage")
grid.arrange(p1,p2,p3,p4)
```


What can we conclude from this? We should conclude that teams need to focus on and prioritize the amount of turnovers they give up, and take away. It is also shown that teams typically need to be offensively dominant in order to win more games.<br>

#### Efficiencies

<br> Next, lets see if we can model overall offensive and defensive efficiencies.I will first look at the standard deviations of each variable to see if there is enough variance for the variable to be impactful.
```{r offensive efficiencies}
off_data = data %>% 
  select(c(1,3:10,12:13,15:18,20,21,24,27))

sds = off_data %>% 
  select(-Team) %>% 
  summarise(across(everything(),sd))
off_data = off_data %>% 
  bind_rows(c(Team = 'SD', sds))
```
<br> Using this method, I see that I should remove the columns of `pass.1st.prop`,`rush.1st.prop`, `int.thrown.pg`, and `t.o.p`.
```{r}
off_data = off_data %>% 
  select(-c(10,14:16))
```


Now, lets proportionalize all of our non-proportional data.<br>
```{r}
sums = off_data %>% 
  select(-Team) %>% 
  summarise(across(everything(),sum))

off_data = off_data %>% 
  bind_rows(c(Team = 'Sum', sums))

off_data = data.frame(t(off_data))
colnames(off_data) = off_data[1,]
off_data = off_data %>% slice(-1)

for (col in 1:ncol(off_data)){
  off_data[[col]] = as.numeric(off_data[[col]])
}

new_df = data.frame(matrix(ncol = ncol(off_data)))
colnames(new_df) = colnames(off_data)
for(row in 1:nrow(off_data)){
  t.row = off_data[row,] / off_data[row,'Sum']
  new_df = rbind(new_df, t.row)
}
new_df = new_df %>% slice(-1) %>% 
  select(-c(SD,Sum))
```
<br> This previous chunk finds the sum of all data, and then divides each number by that sum to find the proportion that each team contributes to this data. This will be the method we use to find offensive efficiencies.<br>
```{r}
totals = new_df %>% 
  summarise(across(everything(),sum))
new_df = new_df %>% 
  bind_rows(totals) 
ranks = new_df %>%
  slice(nrow(new_df)) %>% 
  t() %>% 
  data.frame()
colnames(ranks) = 'rank'
ranks = ranks %>% 
  arrange(-rank)
head(ranks,5)
```
<br> Here, we can see that our top 5 rated offenses are Texas Tech, ISU, Arizona State, Baylor and Colorado.  We think that these are the most efficient offenses, and as already discussed, offensive effiency correlates highly with win percentage. We see this playing out in real life, as 4 of our top 5 offenses finished in the top 5 of the Big 12 Standings. (Texas Tech did not).<br>

### Statistical Leaders

Now, lets look at which teams were best at which categories.
```{r}
ndf = data %>% 
  t() %>% 
  data.frame() 
colnames(ndf) = ndf[1,] 
ndf = ndf %>% slice(-c(1,2))
ndf <- data.frame(lapply(ndf, as.numeric))
nndf = t(apply(ndf, 1, function(x) rank(-x, ties.method = "min")))
rownames(nndf) = colnames(data)[-c(1,2)]
```
<br>This now gives us a data frame that has a teams ranking compared to the other big 12 teams for each statistic, instead of the actual statistic. Next, lets look at the distributions of each teams net rankings.<br>
```{r}
plot_data = melt(nndf)
ggplot(plot_data, aes(x=value))+
  geom_histogram(bins = 16)+
  facet_wrap(~ Var2)+
  labs(title = "Distribution of Category Ranks by Team")+
  xlab('Rank from 1-16')
```
<br>The only significant pattern in any of these histograms is Houston's, which seems to be skewed towards having lower ranks in most categories.It is also important to note that Colorado has a lot of ranks at number 1.<br>

Lets look at this a different way now. 

```{r}
count_rank = plot_data %>% 
  group_by(Var2) %>% 
  summarise(count = sum(value==1)) %>% 
  arrange(-count)
head(count_rank)
ggplot(count_rank, aes(x=reorder(Var2, -count), y = count))+
  geom_col()+
  theme(axis.text.x = element_text(angle = 90))+
  labs(title = "Category Leaders by Count")+
  xlab('')
```
<br> Here we can see the number of categories in which a team leads the big 12 in that specific stat. Colorado and Texas Tech lead in the most categories (5). This is a similar top 5 compared to our offensive efficiency rankings. Note- if there is a tie for first, both teams are credited with leading the big 12.<br>

